---
title: "Wiederholung Tools and Programming"
author: "Alexander Kleine"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: kate
---

```{r knitr_init, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo= T,
	             cache=F,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```


# Lecture 1 - Basics


## Working directory

Please check first where you are by:
getwd()

If you would like to set a new working directory use the following command: 
setwd("/Users/alexander/Documents/Master Kiel/Tools and Programming Languages")


## Simple mathematical operations

```{r}
2+2 # add
5-3 # subtract
10/5 # divide
3*4 # multiply
10^2 # power
sqrt(100) # square root
```


## Assignment operator (<-)

```{r}
a <- 10
b <- 1:5
c <- c('this', 'is', 'a', 'character', 'vector')
d <- list(first_list_object=b, second_list_object=c)
e <- data.frame(column1=b, column2=c)

```


## Inspect your objects

View(e) (Will only be operated in the R-Studio inveroment)
```{r}
rm(a)
rm(b, c)
rm(list=ls())
```


## Inspect your first dataset
```{r}
df <- iris # alternatively write datasets::iris to make the source explicit
head(df) # First 5 observations
class(df) # Verify that it is a dataframe
colnames(df) # Column names
str(df) # Get the structure of the dataframe
levels(df$Species) # Checking the levels of one column
typeof(df) # What is the type of the data frame
plot(df) # Plot the data
attributes(df) # All attributes for each column
```
Why df ist class = data frame and type = list

'mode/type' is a mutually exclusive classification of objects according to their basic structure. The 'atomic' modes are numeric, complex, character and logical. Recursive objects have modes such as 'list' or 'function' or a few others. An object has one and only one mode.

'class' is a property assigned to an object that determines how generic functions operate with it. It is not a mutually exclusive classification. If an object has no specific class assigned to it, such as a simple numeric vector, it's class is usually the same as its mode, by convention.

## Install a package

install.packages('dplyr') (Will only be operated in the R-Studio environment)
```{r}
library('dplyr')
```


## First manipulation of a data frame 

### Different filter functions
```{r}
df2 <- df[df$Sepal.Length>7,]
df3 <- df %>% filter(Sepal.Length>7)
```

## First error and the help function

```{r}
a <- c(1,2,3,4,5,6,7,8,NA)
sum(a)
```

help(sum) (Will only be operated in the R-Studio environment)
Default: sum(..., na.rm = FALSE)
Change the default setting to operate

```{r}
a <- c(1,2,3,4,5,6,7,8,NA)
sum(a, na.rm = TRUE)
```

# Lecture 2 - Basics

### Insert data frame from the package or from the working directory 

df <- read.csv("name.csv")
```{r}
library(datasets)
df <- mtcars
```

## Exploring of the data 

```{r}
class(df)
typeof(df) # how R stores the object
str(df)  # structure
nrow(df)
ncol(df)
dim(df)
head(df)
nrow(df) # How many raws
ncol(df) # How many columns
dim(df) # What is the dimension
tail(x=df, n=5) # Last n rows
head (x=df, n=5) # First n rows
attributes(df)
names(df)
```

## Subsetting of the data

```{r}
df$mpg
df[1:5,]
df[c(1,5),] # c() = combine
df[,1:3]
df[,c("mpg","hp")]
subset <- df[,c("mpg","hp")]
```

## Adding new columns

```{r}
df$new_column <- df$mpg + c(1,100)
df$new_column <- NULL
df$new_column <- df$cyl * df$hp
```


# Lecture 3 - Basics

## Data types
**Vectors** constitute the most important family of data types in R. There are two fundamentally different types of vectors, atomic vectors and lists. **Atomic vectors** have homogeneous element types, i.e. they cannot mix numbers, characters and logical values, whereas **lists** can have heterogeneous element types. 


### Atomic vectors

```{r}
char_scalar <- "data"   # character vector of legth 1
int_scalar <- 1L        # integer vector of length 1
dbl_scalar <- 2.1       # double vector of lenght 1
lgl_scalar <- TRUE      # logical vector of lengh 1

char_scalar
int_scalar
dbl_scalar
lgl_scalar

# The c() function combines multiple smaller vectors into one larger vector

char <- c("data", "science", "rocks")
int <- c(1L, 10L, 100L)
dbl <- c(2.1, pi, 100)
lgl <- c(TRUE, FALSE, FALSE)

char 
int
dbl
lgl
```

***
**Task 1**: Confirm the internal storage type of the atomic vectors `char`, `int`, `dbl`, `lgl` using the function `typeof`? Use the `class`function to check their classes?

```{r}
class(char)
typeof(char_scalar)
```
***

### Type coercion
What happens if we **combine different types** in one vector? R **automatically coerces** to the more flexible type
```{r}
x <- c(char_scalar, int)
x
typeof(x)

```

Not everything is coerced as you might wish: functions may require specific input types:
```{r error=TRUE}
int <- c(1L, 0L, 1L)
char <- c("1", "0", "1")
lgl <- c(TRUE, FALSE, TRUE)
sum(int)
sum(lgl)
sum(char)
```

***
**Task 2**: Rank the 4 common atomic vector types from most / least flexible. To answer this question create different cominations of the atomic vectors `int`, `dbl`, `lgl`, and `char` using the `c()` function and evaluate the type of the combined vector via `typeof()`.
```{r}

```

***


### Factor, date, and datetime objects

There are several other vector classes built on top of atomic vectors. The most important ones are **factor**, **Date** and the datetime object **POSiXct**. They are stored as atomic vectors with attached **attributes**. They all have special properties that are helpful in practice. For instance, the Date class enables us to calculate with dates, sort dates, and print dates in a readable way. 

#### Factors 
**Factors** are built on top of **integer vectors**. They are stored as integers with attached **labels**. Factors are often useful for representing ordered or unordered categorical data.

Let's turn a character vector of clothing sizes sold in a shop into a factor:

```{r}
sizes_char <- c("XXL","S","M","S","L","M", "S", "XXL") # character vector
order <- c("S", "M", "L", "XXL")  # define how clothing sizes are ordered
sizes_factor <- factor(sizes_char, levels=order) # created ordered factor vector
sizes_factor

```

*** 
**Task 3:** Let's try to understand better what a factor is. Apply the functions `class`, `typeof`, `str`, and `attributes` on`sizes_factor` and observe the output.
```{r}
class(sizes_factor)
typeof(sizes_factor)
str(sizes_factor)
attributes(sizes_factor)
```

- `class`: class of the object
- `typeof`: storage mode of the object
- `attributes()`: list of the object' s attributes
- `str()`: compact display of the object's internal structure


**Task 4:** To see why factor variables are sometimes useful, compare the output of the `summary` function for the vectors `sizes_char` and `sizes_factor`. Which output is more useful?
```{r}
summary(sizes_char)
summary(sizes_factor)
```
***

#### Date
The **date** class represents dates as the number of **days since 1970-01-01** and internally stores them as double vector. This enables us to sort dates and calculate with dates: add, subtract, create date sequences, etc.

```{r results='hold'}
today <- Sys.Date()
tomorrow <- today + 1
year_seq <- seq(today, length.out=5, by="1 year")
today
tomorrow
year_seq
```

Let's confirm that 1970-01-01 is day 0 for the R Date class
```{r}
as.integer(as.Date('1970-01-01')) 
```

How many days have passed since day 0?
```{r}
Sys.Date() - as.Date('1970-01-01')
```

***
**Task 5:** How are dates prior to 1970-01-01 stored? Let's try 1969-12-31?
```{r}

```
***

#### Datetime (POSIXct)
The **POSIXct** class represents datetimes as the number of seconds since 1970-01-01 00:00:00 and internally stores them as double vector. This enables us to sort datetimes and calculate with datetimes: add, subtract, create date sequences, etc.

```{r results='hold'}
current_time <- Sys.time()
hour_seq <- seq(current_time, length.out=5, by="10 day")
current_time
hour_seq
```

***
**Task 6**: Check above which timezone is displayed? Is it CEST (central european standard time) or GMT/UTC (Greenwich mean time) or some other time? If you wanted to change your timezone setting, consult `help(timezones)`.

***

### Matrices and arrays
All atomic vectors can be turned into a **matrix** (2-dimensional) or an **array** (multi-dimensional) via a **dimension** attribute. Internally matrices and arrays are stored as atomic vectors, but R treates them differently. If you apply functions on atomic vectors, matrices and arrays, different things will happen.

Let's create an integer vector and convert it into a matrix and an array.

```{r results='hold'}
# Integer vector
int <- 1:12
mat <- matrix(int, nrow=2)
arr <- array(int, dim=c(2,3,2))
``` 

Let's check how matrices and arrays are printed
```{r}
mat  # equivalent to print(mat)
arr  # equivalent to print(arr)
```

Let's analyse the nature of matrices and arrays as compared to the integer vector using the functions `typeof`, `class` and `dim` 
```{r}
# The sapply function applies a function (e.g. typeof) to each element of the provided list of objects.
sapply(X=list(int,mat,arr), FUN=class)   # Class
sapply(list(int,mat,arr), typeof)        # Internal storage
sapply(list(int,mat,arr), dim)           # Dimension
```


***
**Task 7:** Above we turned a vector into a matrix and an array using the dedicated functions `matrix` and `array`. But there's another way to achieve this: by setting the dimension attribute (`dim`). Try some several 2 or 3 dimensional combinations. Observe how printing and the class changes.

```{r}
alphabet <- letters[1:20] # Creates char vector of first 20 letters 
dim(alphabet) <- c(4,5)      # Play around with c(dim1, dim2, [dim3]) 
alphabet                  # Print the object
class(alphabet)           # Display the class
```
***


### List
Lists are **1-dimensional** objects, just like atomic vectors. But different from atomic vectors the list elements can be **heterogeneous.** A list can combine vectors with data frames, arrays and any other R object (functions, formulas, etc.). 

Let's create a simple list containing 3 different vector classes (integer, character, Date) and inspect how the list is printed.

```{r}
simple_list <- list(numbers=1:10, 
                    letters=LETTERS[1:10], 
                    dates=seq(Sys.Date(),length.out =10, by="1 day")
                    )
simple_list
```

Lists are sometimes called recursive vectors, because one can infinitly nest lists within lists. As an example consider the (simplified) representation of the order of Primates.

```{r}
primates <- list(
              name = "Primates",
              Lemures = list(name = "Lemures"),
              Hominoidea = list(
                name = "Hominoidea",
                Orangutans = list(name = "Orang-Utans"),
                Homininae = list(
                  name = "Homininae",
                  Gorillas = list(name = "Gorillas"), 
                  Hominini = list(
                    name = "Hominini",
                    Chimpanzees = list(name = "Chimpanzees"),
                    Humans = list(name = "Humans"))))) 


```

*** 
**Task 8:** To see why nested lists can be useful, install the `data.tree` package and run the code chunk below for a plot of the greatly simplified phylogenetic tree of primates.
```{r message=FALSE, warning=FALSE, error = FALSE}
library(data.tree)
library(data.tree)

plot(as.Node(primates))
```

***

### Data frame
Data frames are **fundamental to data analysis and machine learning**. Data frames are **2-dimensional** like matrices, but can combine **heterogeneous** data types across columns. In terms of structure, a data frame is essentially a **list of equal-length vectors** with attributes for the column names (`names`), row names (`row.names`) and its class (`data.frame`). 

Since the simple list from above consists of equal length vectors we can convert this list into a data frame:
```{r}
df <- data.frame(simple_list) 
df
```

***
**Task 9:** To better understand the nature of data frames apply the functions `class`, `typeof`, `attributes` and `str` on the data frame `df`. 

```{r}
class(df)
typeof(df)
attributes(df)
str(df)
```
**Task 10:** Attributes can be changed. First, change the column names by assigning `c("Zahlen", "Buchstaben", "Daten")` to the `names` attribute. Second, check whether changing the `class` attribute to "list" actually suffices to create a list.
```{r}
df2 <- df %>%
  select(Zahlen = numbers, Buchstaben = letters, Daten = dates)

names(df) <- c("Zahlen", "Buchstaben", "Daten")
```
***

## Generic functions
An important family of R functions is called S3 generic functions. Examples of generic functions are `summary`, `print`, `plot`, and `mean`. Generic functions interact with the class attribute of the functions first argument in a special way. **Depending on the class, generic functions will do different things.**

Let's apply the generic function `summary` to **date**, **factor**, and **character** vectors.

```{r error=FALSE}
summary(year_seq)
summary(sizes_factor)
summary(sizes_char)
```

Do we see that the `summary` function performs three different summary operations for the three different classes of objects? Let's focus on one of the objects, the vector `sizes_factor` and illustrate what the `summary`function is doing under the hood:

1. R checks the `class` of the vector`sizes_factor` 
2. R checks whether there is a dedicated summary method for the Date class (`summary.factor`)
3. **If yes**, the `summary.factor` method is applied. **If no**, the `summary.default` method is applied

```{r}
class(sizes_factor)            # check the class
exists("summary.factor")       # check for factor summary method
summary.factor(sizes_factor)   # apply factor summary method
```

***
**Task 11**: Use the `methods` command to get a list of all methods that can be invoked by the generic functions `summary` and `mean`.
```{r}
methods(mean)
methods(summary)
```
***


## Selecting parts of an object
We often wish to **select parts of an object for the purpose of extraction or replacement**.  We have mainly three operators to chose , `[`, `[[` and `$`.

### Multiple elements
The operator `[` allows **extracting** any element or any combination of elements of an R object. Within the square brackets we specify along which dimensions we want to select, writing `[dim1, dim2, ...]`. In case of matrices and data frames this is `[row, col]`.

In general, the dimension arguments within the square brackets can take three different forms: 

1. Numeric vector
2. Logical vector
3. Character vector

#### Atomic vector
```{r results='hold'}
int <- 1:5
names(int) <- letters[1:5]              # An atomic vector can be named
int
```

**Extracting**
```{r results='hold'}
int[1]                                  # R uses 1-based indexing
int[2:3]                                  # R uses 1-based indexing
int[c(1,3)]                             # numeric 
int[c(TRUE, FALSE, TRUE, FALSE, FALSE)] # logical 
int[c("a", "c")]                        # character (matched to names)
```

**Replacing**

Note that the dimensions have to match!
```{r error=FALSE}
int[c(1,3)]  <- c(100, 300, 500) # dimensions don't match, hence we get an error
```

```{r}
int[c(1,3)]  <- c(100, 300) # here the dimensions match
int

```
#### Matrix
```{r}
mat <- matrix(1:12, nrow = 3, ncol=4)
rownames(mat) <- paste0("r", 1:3)    # matrices can have named dimensions
colnames(mat) <- paste0("c", 1:4)  
mat

mat[c(1,2), c(3,4)]                                    # numeric
mat[c(TRUE, TRUE, FALSE), c(FALSE, FALSE, TRUE, TRUE)] # logical
mat[c("r1", "r2"), c("c3", "c4")]                      # character
mat[c(1,2), ]                                          # leave blank dim2
```



### Special cases
#### Negative selection
We can invert a selection using the `-` operator for **numeric vectors** and using the `!` operator for **logical vectors**.
```{r}
int[c(1, 3)]                              # numeric
int[-c(1, 3)]                             # numeric
int[c(TRUE, FALSE, TRUE, FALSE, FALSE)]   # logical
int[!c(TRUE, FALSE, TRUE, FALSE, FALSE)]  # logical
```


### Single elements
There are two other important operators: `[[` and `$`. They are usful in the **context of lists and data frames** (which are internally stored as lists) when we want to **select only 1 element**, e.g. one column of a data frame. While the `[` operator preserves the list structure, `[[` and `$` enable us to **navigate into the list structure**

Let's consider our simple list from above:
```{r results='hold'}
simple_list["dates"]           # preserves the list class
simple_list[["dates"]]         # extracts a Date vector
simple_list$dates              # also extracts a Date vector
```

The advantage becomes aparent for nested lists. Different from `[`, the operators ``[[`` and `$` allow **navigating deeply into a nested list**, and extract or replace elements there.

***
**Task 12**: Use the $ operator multiple times to navigate through the nested list `primates` until you reach `humans`. Don't type everything by hand. Use autocompletion by pressing `tab` after each `$` operator.

```{r}
primates$Hominoidea$Homininae$Hominini$Humans$name
  
```

Given that `$` is less verbose and easier to read, **why and when should we use `[[` instead**? Answer: In situations where we want to pass the selection as a variable.
```{r results='hold'}
selected_element <- "numbers"

str(simple_list)
simple_list[[selected_element]] # works fine
simple_list$selected_element    # does not work
```


### Application

#### Data frame
Now, let's focus on dataframes and on some applications of extraction and replacement that we often see in practice. But note that we will cover packages like `dplyr` or `data.table` later in class, which make subsetting data frames much more convenient. 

```{r}
x <- 1:20
condition <- x>5 # The condition x>5 creates a logical index
condition
x[condition]     # We subset vector x using this index
x[!condition]    # We implement negative selection
```
***

**Task 13:** The iris data frame contains information for 150 flowers. *First*, extract all 12 observations with sepal length larger than 7. *Second*, for these flowers only select the `Species` column. *Third*, change the species names of all flowers to capital letters using the `toupper` function. *Fourth*, create an additional variable `sepal_length_100` by multiplying `Sepal.Length` by factor 100.

```{r}
df <- iris

# We explicitly coerce from factor to character. Otherwise toupper won't work.
df$Species <- as.character(df$Species) 

# First
condition <- df$Sepal.Length > 7 

df1 <- df[condition, ]

condition2 <- df1$Petal.Width == 2.1

df1.2 <- df1[condition2,]

df1.2
df1


# Second

df2 <- df[condition,"Species"]

# Third

df3 <- toupper(df$Species)

#Fourth
df4 <- df
df4$sepal_length_100 <- df4$Sepal.Length*100
```


**Task 14:** In supervised machine learning its common practice to split the data randomly into a training set (70% of the observations) and a test set (30%). Perform this split for the iris data frame, and use *negative selection* to create the test data.

```{r}
# Use the following index for this task
N <- nrow(df)
index <- sample(N, size=N * 0.7)

train <- df[index,]
traintest <- sample_n(df, nrow(df)*0.7 )

test <- df[-index,]
testtest <- sample_n(df, nrow(df)*0.3 )

```


## Loops and their alternatives
Instead of using loops it is often more R-like to use `apply` (base R) or `map` (purrr package). However, using a vectorized function like `+` or `*`, if available, is the preferred solution because the code is more performant.

As an example consider the following `simple_vector`. First, we want to multiply each element of this vector by factor 2 using a `for` loop:
```{r}
simple_vector <- 1:100

doubled_vector <- simple_vector # First we initialize the doubled_vector
for (x in simple_vector){
  doubled_vector[x] <- x*2      # Then we iterate through each element and multiply by 2 each time
}
doubled_vector
```

Now, let's reformulate this operation using a member of the `apply` family of functions:
```{r}
# Here we make use of an anonymous function.
# Since it is anonymous it cannot be used elsewhere

doubled_vector <- sapply(simple_vector, function(x) x*2) 

```

***
**Task:** Reformulate this operation using the vectorized `*` function. This is the most efficient and syntactically easiest way to do it:
```{r}

```

## Get data in and out

***
**Task:** Use one of the data frames of the `datasets` package to experiment with importing and exporting. And to practice some of the other aspects of base R, covered before. Experiment with subsetting, extract or change columns of the data frame, inspect or change (coerce) the data types of columns, etc. Then you can write to `.csv`, `.RData` and `.Rds`. Clear the environment and import the data back again. 
```{r}


```

?lm

